# Added variable plots

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,
                      warning = FALSE,
                      cache = TRUE)
```

p. 54 Figure 3.3(a)

1. Compute the regression of `lifeExpF` on `log(ppgdp)`:

```{r}
library(alr4)
library(tidyverse)
library(plotly)
u <- UN11
```


```{r}
mod1 <- lm(u$lifeExpF ~ log(u$ppgdp))
mod1
summary(mod1)$r.squared
u$`Model 1 residuals` <- ifelse(mod1$residuals > 0, "higher than expected lifeExpF", "lower than expected lifeExpF")

```

```{r}
g <- ggplot(u, aes(log(ppgdp), lifeExpF, label = rownames(u), color = `Model 1 residuals`)) + 
  geom_point() +  
  geom_smooth(aes(group = 1), method = "lm", se=FALSE, show.legend =FALSE) +
  ggtitle("Model 1")
g

```


**The residuals from this model are the part of the response `lifeExpF` not explained by the regression on `log(ppgdp)`**


2. Compute the regression of `fertility` on `log(ppgdp)`.

**The residuals from this model are the part of the new regressor `fertility` not explained by the regression on `log(ppgdp)`**
```{r}
mod2 <- lm(u$fertility ~ log(u$ppgdp))
mod2
summary(mod2)$r.squared

g <- ggplot(u, aes(log(ppgdp), fertility, label = rownames(u), color = `Model 1 residuals`)) + 
  geom_point() +
  geom_smooth(aes(group = 1), method = "lm", se=FALSE, show.legend = FALSE) +
  ggtitle("Model 2")
g
```

3. The added variable plot is of the unexplained part of the response from (1) on the unexplained part of the added regressor from (2).

```{r}
residdf <- tibble::tibble(r1 = mod1$residuals,
                          r2 = mod2$residuals)
g <- ggplot(residdf, aes(r2, r1)) + 
  geom_point() +
  xlab(expression(paste(hat(e), " from fertility on log(ppgdp)"))) +
  ylab(expression(paste(hat(e), " from lifeExpF on log(ppgdp)"))) +
  geom_smooth(method = "lm", se=FALSE)
g
```
The slope of this regression line is:

```{r}
lm(residdf$r1~residdf$r2)
```
It is the same as the coefficient of $\hat{\beta_2}$ with both regressors:

```{r}
u$logppgdp <- log(u$ppgdp)
mod <- lm(lifeExpF ~ fertility + logppgdp, data = u)
mod
```

```{r, eval=FALSE, echo=FALSE}
# how to label axes in base R
plot(mod1$residuals ~ mod2$residuals, pch = 16, cex = .75, xlab = expression(paste(hat(e), " from fertility on log(ppgdp)")), ylab = expression(paste(hat(e), " from lifeExpF on log(ppgdp)")), las = 1)
abline(lm(mod1$residuals ~ mod2$residuals))

# how to label axes in plotly
ggplotly(g) %>%  layout(
    xaxis = list(title = TeX("\\hat{e} \\text{ from fertility on log(ppgdp)}")),
    yaxis = list(title = TeX("\\hat{e} \\text{ from lifeExpF on log(ppgdp)}"))) %>%
  config(mathjax = "cdn")
```


